Global patterns of forest autotrophic carbon fluxes

### 2019-09-12 manuscript/manuscript draft.Rmd
fixed some math syntex

| Variable | Definition | Components included | Methodologies used | Number of records | Number of geographic areas |
|--------------|-------------------|-------------|------------------------|-------|-------|
| GPP | Annual gross primary production; annual uptake of carbon dioxide by an ecosystem | NA | Flux partitioning of eddy covariance | 243 | 49 |
| NPP | Annual net primary production; the component of GPP that is stored in plant tissue; GPP minus ecosystem respiration | Foliage, branch, stem, coarse root and fine root | Direct measurement of annual increments of components | 92 | 42 |
| ANPP | Aboveground net primary production | Foliage, stem, and optionally branch | Direct measurement of annual increments of components | 256 | 80 |
| ANPP~foliage~ | Net primary production of foliage | Foliage | Direct measurement of litterfall, correcting for changes in leaf biomass when measured | 98 | 49 |
| ANPP~woody~ ~stem~ | Net primary production of woody stems | Woody stems | Direct measurement of stem growth increment | 264 | 96 |
| BNPP~root~ | Belowground net primary production | Coarse and fine roots | Direct measurement of one or more of: fine root turnover, soil cores, root ingrowth cores, minirhizotrons; indirect estimates of coarse roots using allometries based on aboveground stem increment measures | 101 | 48 |
| BNPP~fine~ ~root~ | Net primary production of fine roots | Fine roots | Direct measurement of one or more of: minirhizotrons, fine root turnover, soil cores, root ingrowth cores | 88 | 41 |
| R~auto~ | Annual autotrophic respiration, including above- and belowground components | Foliage, stem, and root | Chamber measurements of component gas exchange | 22 | 13 |
| R~root~ | Annual root respiration | Coarse and fine roots | Measurement of root gas exchange | 64 | 26 |

@@

As discussed above, estimates of NPP, ANPP, and BNPP are generated through summing measurements of their component parts. Since the components included in productivity estimates vary between studies, estimates of productivity were classified within the ForC database according to their components, and then filtered for analysis. Estimates of NPP were selected if they included foliage, branch, stem, coarse root, and fine root. Measures of NPP which included additional components, including understorey, volatile organic compounds (VOCs), exudates, estimates of NPP lost to herbivory, and the NPP of reproductive structures, were excluded. Estimates of ANPP were selected if they included foliage, stem growth and optionally branch turnover. Any measures of primary productivity where components were unknown were excluded from analysis.

@@

*Model specification.* The effects of climate and latitude on primary productivity were analysed using mixed effects models using the package 'lme4' [@bates_fitting_2015] in R v.3.5.1 [@r_core_team_r:_2018]. The effect of each extracted climate variable on each measure of primary productivity was modelled by specifying the climate variable as a fixed effect. For each climate variable, three models were specified: a null model; a model with the climate variable as a linear term; and a model with the climate variable as a polynomial term. AIC values were calculated for the models and used to select the best model. If the best model included a polynomial term, the shape of the polynomial relationship was considered. If the shape of the relationship *made biological sense*, and was a significant improvement on the linear relationship (deltaAIC >2), we accepted the polynomial as the best model. If not, we ran the linear model as the final model. R^2^ values were calculated for the best model. All R^2^ values presented here are marginal R^2^ values, and refer to the proportion of variation explained by only the fixed effects, unless otherwise specified. In addition, slope coefficients were calculated for the linear models.

@@

R^2^ values were generally highest in the major fluxes, and decreased in subsidiary fluxes (Supporting Information S2). Of the major fluxes, R~auto~ and GPP were the most strongly explained by latitude and climate, with climate explaining at most 71% of variation in GPP, and 65% in R~auto~. The proportion of variation explained by climate and latitude decreased in NPP and ANPP, with climate explaining at most 51% of variation in NPP and 44% in ANPP. Of the major fluxes, BNPP~root~ was the least well explained by climate and latitude, with climate explaining at most 36% of variation.

With the exception of ANPP~foliage~, the proportion of variation explained by climate and latitude in subsidiary fluxes was much lower. Climate explained at most 24% of variation in ANPP~woody~ ~stem~, 19% in BNPP~fine~ ~root~, and 27% in R~root~. In contrast, climate strongly explained variation in ANPP~foliage~, with mean annual temperature explaining 58% of variation. This pattern was also seen in the R^2^ values for multivariate models.

@@

MAP was found to be a significant (p<0.05) but poor predictor of productivity, explaining, with the exception of R~auto~, at most 37% of variation in carbon flux. For the majority of fluxes productivity was best predicted by a polynomial model. Productivity increased with precipitation, up until a saturation point at between 3000 and 4000mm annual precipitation, above which productivity started to decrease (fig. 5). The notable exception to this was GPP: the model indicated that GPP continued to increase with precipitation up to measures of at least 5000mm annually (p<0.0001, R^2^ = 0.33. Data above this point was not available, but the model trend indicated that the saturation point for this model would be around 5000mm MAP.

@@

Vapour pressure deficit was a significant predictor of productivity for all fluxes. BNPP~fine~ ~root~ showed a linear relationship with vapour pressure deficit (R^2^ = 0.07, p<0.05), but all other fluxes showed a polynomial relationship (fig. 5). Productivity initially increased with vapour pressure deficit, before saturating at around 0.8 kPa. At values above 0.8 kPa, productivity began to decrease.

All fluxes, with the exception of R~root~, showed a positive linear relationship with solar radiation. Solar radiation explained a low proportion of variability in productivity for all fluxes, explaining less than 20% of the variation in each flux, with the exception of R~auto~ (R^2^ = 0.26, p<0.05).

@@

*Within the growing season, how does productivity vary with climate?* Within growing season months, we found that climate had a much weaker effect on productivity. For each of temperature, precipitation, PET, and solar radiation, we found a small effect of climate for certain carbon fluxes. There was a small increase in productivity with temperature and precipitation for ANPP (with temperature R^2^ = 0.10, p<0.001; with precipitation R^2^ = 0.04, p<0.05) and ANPP~foliage~ (with temperature R^2^ = 0.16, p<0.01; with precipitation R^2^ = 0.09, p<0.05). Productivity increased with solar radiation for GPP (R^2^ = 0.21, p<0.001), NPP (R^2^ = 0.21, p<0.001), BNPP (R^2^ = 0.16, p<0.001) and BNPP~fine~ ~root~ (R^2^ = 0.12, p<0.01), and with PET for GPP (R^2^ = 0.15, p<0.01), NPP (R^2^ = 0.18, p<0.01), BNPP (R^2^ = 0.23, p<0.0001), BNPP~fine~ ~root~ (R^2^ = 0.11, p<0.05), and ANPP~woody~ ~stem~ (R^2^ = 0.06, p<0.05).

@@

Past studies have differed in their conclusions regarding the relationship between productivity and latitude or its correlates (**REFS**)--quite possibly because of lack of standardization with respect to stand age and disturbance history. Our findings indicate that, among mature, undisturbed stands, FACF is unambiguously highest in the tropical regions, and the relationship is approximately linear (Figs. 2-3). This contrasts with suggestions that productivity of temperate forests is similar to that of tropical forests (**REFS**). Such observations may be attributable to the fact that temperate forests are predominantly secondary and of younger age than tropical forests (**REF**), so analyses comparing across latitudinal gradients without strict control of stand age tend to be comparing younger temperate stands with older tropical (and often boreal) forests ([Poulter et al. 2018-DOI:10.1594/PANGAEA.889943](http://pure.iiasa.ac.at/id/eprint/15361/)). Because C allocation varies with stand age (**REFS (KAT/NOBBY)**), age differences may introduce systematic biases into analyses of FACF across latitude or global climatic gradients. For example, woody productivity tends to be higher in rapidly aggrading secondary stands than in old-growth forests, where proportionally more C is allocated to respiration (**REFS**) [*purpose for respiration/ other compenents (**REFS????- NOBBY?**)]. By constraining our analysis to stands $\\geq$100 years old with no record of major recent disturbance, we have clarified the shape of the relationship between FACF and latitude.

@@

One interesting observation was that climate tends to explain more variation in the major fluxes ($GPP$, $NPP$, $R_{auto}$ - latitude $R^2 \ge$48%) than in subsidiary fluxes (latitude $R^2$<27% for $BNPP_{fine.root}$, $R_{auto-root}$,$ANPP_{woody-stem}$) (Fig. 2; **some SI table?**). There are two non-exclusive potential explanations for this. First, it may be that methodological variation is larger relative to flux magnitude for some of the subsidiary fluxes. Belowground fluxes in particular are difficult to quantify, and measurement methods for the belowground fluxes considered here may be measured through fundamentally different approaches (*e.g.*, minirhizotrons, ingrowth cores, or sequential coring for $BNPP_{root-fine}$; root exclusion, stable isotope tracking, or gas exchange of excised roots for $R_{auto-root}$), and sampling depth is variable and often insufficient to capture the full soil profile. $ANPP_{woody-stem}$, which is also poorly explained by latitude or climate, is more straightforward to measure but is subject to variability introduced by differences such as biomass allometries applied and minimum plant size sampled. However, methodological variation and uncertainty affect all of fluxes considered here--not necessarily any less than the aforementioned, and some of the larger fluxes that vary more strongly with respect to climate ($ANPP$, $NPP$) are estimated by summing uncertain component fluxes. Second, differences among variables in the proportion of variation explained by climate may be attributable to more dicrect climatic control over $GPP$ than subsidiary fluxes. That is, subsidiary fluxes may be shaped by climate both indirectly through its influence on $GPP$ and potentially respiration and directly through any climatic influence on C allocation, as well as by many other local- and regional-scale factors (**REFS**).

>>>>

| Variable | Definition | Components included | Methodologies used | Number of records | Number of geographic areas |
|--------------|-------------------|-------------|------------------------|-------|-------|
| $GPP$ | Annual gross primary production; annual uptake of carbon dioxide by an ecosystem | NA | Flux partitioning of eddy covariance | 243 | 49 |
| $NPP$ | Annual net primary production; the component of GPP that is stored in plant tissue; GPP minus ecosystem respiration | Foliage, branch, stem, coarse root and fine root | Direct measurement of annual increments of components | 92 | 42 |
| $ANPP$ | Aboveground net primary production | Foliage, stem, and optionally branch | Direct measurement of annual increments of components | 256 | 80 |
| $ANPP_{foliage}$ | Net primary production of foliage | Foliage | Direct measurement of litterfall, correcting for changes in leaf biomass when measured | 98 | 49 |
| $ANPP_{woody-stem}$ | Net primary production of woody stems | Woody stems | Direct measurement of stem growth increment | 264 | 96 |
| $BNPP_{root}$  | Belowground net primary production | Coarse and fine roots | Direct measurement of one or more of: fine root turnover, soil cores, root ingrowth cores, minirhizotrons; indirect estimates of coarse roots using allometries based on aboveground stem increment measures | 101 | 48 |
| $BNPP_{fine.root}$ | Net primary production of fine roots | Fine roots | Direct measurement of one or more of: minirhizotrons, fine root turnover, soil cores, root ingrowth cores | 88 | 41 |
| $R_{auto}$ | Annual autotrophic respiration, including above- and belowground components | Foliage, stem, and root | Chamber measurements of component gas exchange | 22 | 13 |
| $R_{root}$ | Annual root respiration | Coarse and fine roots | Measurement of root gas exchange | 64 | 26 |

@@

As discussed above, estimates of $NPP$, $ANPP$, and $BNPP$ are generated through summing measurements of their component parts. Since the components included in productivity estimates vary between studies, estimates of productivity were classified within the ForC database according to their components, and then filtered for analysis. Estimates of NPP were selected if they included foliage, branch, stem, coarse root, and fine root. Measures of NPP which included additional components, including understorey, volatile organic compounds (VOCs), exudates, estimates of NPP lost to herbivory, and the NPP of reproductive structures, were excluded. Estimates of ANPP were selected if they included foliage, stem growth and optionally branch turnover. Any measures of primary productivity where components were unknown were excluded from analysis.

@@

*Model specification.* The effects of climate and latitude on primary productivity were analysed using mixed effects models using the package 'lme4' [@bates_fitting_2015] in R v.3.5.1 [@r_core_team_r:_2018]. The effect of each extracted climate variable on each measure of primary productivity was modelled by specifying the climate variable as a fixed effect. For each climate variable, three models were specified: a null model; a model with the climate variable as a linear term; and a model with the climate variable as a polynomial term. AIC values were calculated for the models and used to select the best model. If the best model included a polynomial term, the shape of the polynomial relationship was considered. If the shape of the relationship *made biological sense*, and was a significant improvement on the linear relationship (deltaAIC >2), we accepted the polynomial as the best model. If not, we ran the linear model as the final model. $R^2$ values were calculated for the best model. All $R^2$ values presented here are marginal $R^2$ values, and refer to the proportion of variation explained by only the fixed effects, unless otherwise specified. In addition, slope coefficients were calculated for the linear models.

@@

$R^2$ values were generally highest in the major fluxes, and decreased in subsidiary fluxes (Supporting Information S2). Of the major fluxes, R~auto~ and GPP were the most strongly explained by latitude and climate, with climate explaining at most 71% of variation in GPP, and 65% in R~auto~. The proportion of variation explained by climate and latitude decreased in NPP and ANPP, with climate explaining at most 51% of variation in NPP and 44% in ANPP. Of the major fluxes, BNPP~root~ was the least well explained by climate and latitude, with climate explaining at most 36% of variation.

With the exception of ANPP~foliage~, the proportion of variation explained by climate and latitude in subsidiary fluxes was much lower. Climate explained at most 24% of variation in ANPP~woody~ ~stem~, 19% in BNPP~fine~ ~root~, and 27% in R~root~. In contrast, climate strongly explained variation in ANPP~foliage~, with mean annual temperature explaining 58% of variation. This pattern was also seen in the $R^2$ values for multivariate models.

@@

MAP was found to be a significant (p<0.05) but poor predictor of productivity, explaining, with the exception of R~auto~, at most 37% of variation in carbon flux. For the majority of fluxes productivity was best predicted by a polynomial model. Productivity increased with precipitation, up until a saturation point at between 3000 and 4000mm annual precipitation, above which productivity started to decrease (fig. 5). The notable exception to this was GPP: the model indicated that GPP continued to increase with precipitation up to measures of at least 5000mm annually (p<0.0001, $R^2$ = 0.33. Data above this point was not available, but the model trend indicated that the saturation point for this model would be around 5000mm MAP.

@@

Vapour pressure deficit was a significant predictor of productivity for all fluxes. BNPP~fine~ ~root~ showed a linear relationship with vapour pressure deficit ($R^2$ = 0.07, p<0.05), but all other fluxes showed a polynomial relationship (fig. 5). Productivity initially increased with vapour pressure deficit, before saturating at around 0.8 kPa. At values above 0.8 kPa, productivity began to decrease.

All fluxes, with the exception of R~root~, showed a positive linear relationship with solar radiation. Solar radiation explained a low proportion of variability in productivity for all fluxes, explaining less than 20% of the variation in each flux, with the exception of R~auto~ ($R^2$ = 0.26, p<0.05).

@@

*Within the growing season, how does productivity vary with climate?* Within growing season months, we found that climate had a much weaker effect on productivity. For each of temperature, precipitation, PET, and solar radiation, we found a small effect of climate for certain carbon fluxes. There was a small increase in productivity with temperature and precipitation for ANPP (with temperature $R^2$ = 0.10, p<0.001; with precipitation $R^2$ = 0.04, p<0.05) and ANPP~foliage~ (with temperature $R^2$ = 0.16, p<0.01; with precipitation $R^2$ = 0.09, p<0.05). Productivity increased with solar radiation for GPP ($R^2$ = 0.21, p<0.001), NPP ($R^2$ = 0.21, p<0.001), BNPP ($R^2$ = 0.16, p<0.001) and BNPP~fine~ ~root~ ($R^2$ = 0.12, p<0.01), and with PET for GPP ($R^2$ = 0.15, p<0.01), NPP ($R^2$ = 0.18, p<0.01), BNPP ($R^2$ = 0.23, p<0.0001), BNPP~fine~ ~root~ ($R^2$ = 0.11, p<0.05), and ANPP~woody~ ~stem~ ($R^2$ = 0.06, p<0.05).

@@

Past studies have differed in their conclusions regarding the relationship between productivity and latitude or its correlates (**REFS**)--quite possibly because of lack of standardization with respect to stand age and disturbance history. Our findings indicate that, among mature, undisturbed stands, FACF is unambiguously highest in the tropical regions, and the relationship is approximately linear (Figs. 2-3). This contrasts with suggestions that productivity of temperate forests is similar to that of tropical forests (**REFS**). Such observations may be attributable to the fact that temperate forests are predominantly secondary and of younger age than tropical forests (**REF**), so analyses comparing across latitudinal gradients without strict control of stand age tend to be comparing younger temperate stands with older tropical (and often boreal) forests ([Poulter et al. 2018-DOI:10.1594/PANGAEA.889943](http://pure.iiasa.ac.at/id/eprint/15361/)). Because C allocation varies with stand age (**REFS (KAT/NOBBY)**), age differences may introduce systematic biases into analyses of FACF across latitude or global climatic gradients. For example, woody productivity tends to be higher in rapidly aggrading secondary stands than in old-growth forests, where proportionally more C is allocated to respiration (**REFS**) [*purpose for respiration/ other compenents (**REFS????- NOBBY?**)]. By constraining our analysis to stands $\ge$ 100 years old with no record of major recent disturbance, we have clarified the shape of the relationship between FACF and latitude.

@@

One interesting observation was that climate tends to explain more variation in the major fluxes ($GPP$, $NPP$, $R_{auto}$ - latitude $R^2\ge$ 48%) than in subsidiary fluxes (latitude $R^2$ <27% for $BNPP_{fine.root}$, $R_{auto-root}$,$ANPP_{woody-stem}$) (Fig. 2; **some SI table?**). There are two non-exclusive potential explanations for this. First, it may be that methodological variation is larger relative to flux magnitude for some of the subsidiary fluxes. Belowground fluxes in particular are difficult to quantify, and measurement methods for the belowground fluxes considered here may be measured through fundamentally different approaches (*e.g.*, minirhizotrons, ingrowth cores, or sequential coring for $BNPP_{root-fine}$; root exclusion, stable isotope tracking, or gas exchange of excised roots for $R_{auto-root}$), and sampling depth is variable and often insufficient to capture the full soil profile. $ANPP_{woody-stem}$, which is also poorly explained by latitude or climate, is more straightforward to measure but is subject to variability introduced by differences such as biomass allometries applied and minimum plant size sampled. However, methodological variation and uncertainty affect all of fluxes considered here--not necessarily any less than the aforementioned, and some of the larger fluxes that vary more strongly with respect to climate ($ANPP$, $NPP$) are estimated by summing uncertain component fluxes. Second, differences among variables in the proportion of variation explained by climate may be attributable to more dicrect climatic control over $GPP$ than subsidiary fluxes. That is, subsidiary fluxes may be shaped by climate both indirectly through its influence on $GPP$ and potentially respiration and directly through any climatic influence on C allocation, as well as by many other local- and regional-scale factors (**REFS**).
